---
title: "1. Geographic Assignment"
subtitle: "Using the assignR package to estimate geographic origin"
description: "Shown in search results and link previews."
date-modified: last-modified
---

::: {.callout-note appearance="simple" icon="false"}
## In this section

-   **You'll learn:** the key idea or skill
-   **You'll need:** packages or data from earlier sections
-   **Time:** about 30 minutes
:::

## Overview

A short introduction to the topic: why it matters and what we're going to do.

## Setting up

```{r}
#| label: setup
library(tidyverse)
library(assignR)
library(raster)
```

## Unknown origin data

Explain the idea in a sentence or two, then show the code.

```{r}
#| label: first-example
mtcars |>
  group_by(cyl) |>
  summarise(mean_mpg = mean(mpg))
```

::: callout-tip
## Tip

A helpful shortcut or good-practice note.
:::

### A sub-step

Use `###` headings for smaller steps. They appear nested in the table of contents.

## Strontium isoscape

Tabsets are handy for showing alternatives side by side:

::: panel-tabset
## tidyverse

```{r}
#| label: tidy-version
mtcars |> filter(mpg > 25)
```

## base R

```{r}
#| label: base-version
mtcars[mtcars$mpg > 25, ]
```
:::

::: callout-warning
## Common mistake

Describe an error students often hit and how to fix it.
:::

## Isotope-based geographic assignment

Tabsets are handy for showing alternatives side by side:

## Exercise

::: {.callout-tip icon="false"}
## 🧩 Your turn

1.  First task

2.  Second task

3.  **Stretch:** an extra challenge for students who finish early

    ```{r}


    ```
:::

::: {.callout-caution collapse="true"}
## Hint

A nudge in the right direction, hidden until clicked.
:::

::: {.callout-caution collapse="true"}
## Solution

```{r}
#| label: exercise-solution
#| eval: false
# Solution code here
```
:::

## Summary

::: {.callout-note appearance="simple"}
## Key takeaways

-   First key point
-   Second key point
:::

**Next:** [Section 2 →](02-next-section.qmd)
