---
title: "2. Combining geolocation tools"
subtitle: "multiple isotopes and priors"
description: "multiple isotopes and priors"
date-modified: last-modified
---

::: {.callout-note appearance="simple" icon="false"}
## In this section

-   **You'll learn:** the key idea or skill
-   **You'll need:** packages or data from earlier sections
-   **Time:** about 30 minutes
:::

## Overview

Two are better than one! Many studies have shown that multi-isotope assignment can result in better spatial precision than single-isotope assignment. Let's try it!

## Setting up

```{r}
#| label: setup
library(tidyverse)
library(assignR)
library(raster)
```

### Hydrogen isoscape

Hydrogen isotopes are commonly used for geographic assignment of butterflies. Spatial variation in hydrogen isotopes is largely driven by patterns of precipitation. Since this pattern is independent of strontium isotopes, the two isotopes are great complementary geolocation tools.

How does the hydrogen isoscape compare to the strontium isoscape?

```{r}
d2Hw_isoscape <- terra::rast("Module3_assignment/d2Hw_isoscape.tif")

# how does the spatial pattern compare to strontium isotopes?
plot(d2Hw_isoscape); plot(Sr_iso)
```
What do you notice about the patterns in the mean and sd layers of the hydrogen isoscape? Where are the predicted d2H values high? why? where are they low? Is there a correlation between predicted d2H values and uncertainty? Where is uncertainty high? low?

::: callout-warning
## Careful!

It is important to use a hydrogen isoscape that has been calibrated to the material of interest. Here, I am giving you a hydrogen isoscape that has already been calibrated to insect wings. Calibrating a hydrogen isoscape involves (almost) as many decisions as developing a strontium isoscape.
:::

## Hydrogen isotope-based geographic assignment
What do the hydrogen-only probability surfaces look like?
```{r}
unknown_H <- unknown %>%
  dplyr::select(Sample_ID, d2H, d2H_sd) #RVcoll07F553 is above range of South Africa isoscape

H_probsurfaces <- assignR::pdRaster(r = d2Hw_isoscape, unknown = unknown_H)
```


### Preparing for dual assignment

For simplicity, we will append d2H values for each butterfly and assume the measurement uncertainty is 2 permille.

```{r}
# arrange dataframe to assignR specifications
unknown_dual <- unknown %>%
  dplyr::select(Sample_ID, Sr8786, d2H)

dual_isoscapes <- isoStack(Sr_iso_crop, d2Hw_isoscape)

```

### dual assignment

```{r}
# hydrogen and strontium isotope-based geographic assignment
dual_assign <- pdRaster(r = dual_isoscapes, unknown = unknown_dual)
```

How do these posterior probability surfaces compare to the ones generated with strontium isotopes alone? Is there a smaller area of high probability? ::: panel-tabset \## tidyverse

```{r}
#| label: tidy-version
#plot(dual)
```

## base R

```{r}
#| label: base-version
#plot(Sr_assign)
```

## Using a prior

::: {.think-pair-share}
### 🦋 Think · Pair · Share

::: {.tps-steps}
::: {.tps-step}
**💭 Think** <span class="tps-time">1 min</span>

On your own: what do you already know about where *your* study species could
have come from, before measuring any strontium?
:::

::: {.tps-step}
**🤝 Pair** <span class="tps-time">2 min</span>

Turn to your neighbour. How could you turn that knowledge into a prior map,
P(j)? Which cells would get a prior of zero?
:::

::: {.tps-step}
**📣 Share** <span class="tps-time">2 min</span>

Tell the room one idea for a prior, and one way it could go wrong.
:::
:::
:::

### Ecological niche model



::: {.callout-caution collapse="true"}
## Hint

A nudge in the right direction, hidden until clicked.
:::

::: {.callout-caution collapse="true"}
## Solution

```{r}
#| label: exercise-solution
#| eval: false
# Solution code here
```
:::

## Summary

::: {.callout-note appearance="simple"}
## Key takeaways

-   Combining multiple geolocation tools can give spatially constrained estimates of origin
-   Each geolocation tool comes with additional assumptions and uncertainty
:::

```{r}

```

**Next:** [Section 3 →](03-section-summarising.qmd)
