---
title: "3. Summarizing outputs"
subtitle: "Summarize posterior probability surfaces"
description: "Summarize posterior probability surfaces"
date-modified: last-modified
---

::: {.callout-note appearance="simple" icon="false"}
## In this section

-   **You'll learn:** the key idea or skill
-   **You'll need:** packages or data from earlier sections
-   **Time:** about 30 minutes
:::

## Overview

A short introduction to the topic: why it matters and what we're going to do.

## Setting up

```{r}
#| label: setup
library(tidyverse)
library(assignR)
library(raster)
library(isocat)
```

## Odds ratio

Explain the idea in a sentence or two, then show the code.

```{r}
#| label: first-example
mtcars |>
  group_by(cyl) |>
  summarise(mean_mpg = mean(mpg))
```

::: callout-tip
## Tip

A helpful shortcut or good-practice note.
:::

## Cluster analysis

## Spatial similarity with isocat

For simplicity, we will append d2H values for each butterfly and assume the measurement uncertainty is 2 permille.

```{r}
# append d2H values for each butterfly
#d2H <- c( )
#d2H.sd <- c()

# arrange dataframe to assignR specifications


```

```{r}
#| label: tidy-version
#plot(dual)
```

::: {.callout-tip icon="false"}
## 🧩 Your turn

1.  First task
2.  Second task
3.  **Stretch:** an extra challenge for students who finish early
:::

::: {.callout-caution collapse="true"}
## Hint

A nudge in the right direction, hidden until clicked.
:::

::: {.callout-caution collapse="true"}
## Solution

```{r}
#| label: exercise-solution
#| eval: false
# Solution code here
```
:::

## Summary

::: {.callout-note appearance="simple"}
## Key takeaways

-   Combining multiple geolocation tools can give spatially constrained estimates of origin
-   Each geolocation tool comes with additional assumptions and uncertainty
:::

```{r}
#**Next:** [Section 2 →](02-next-section.qmd)
```
