---
title: "3. Summarizing outputs"
subtitle: "Summarize posterior probability surfaces"
date-modified: last-modified
---

::: {.callout-note appearance="simple" icon="false"}
## In this section

-   **You'll learn:** a few ways to summarize your results
-   **You'll need:** the same packages and data from earlier sections
-   **Time:** about 30 minutes
:::

## Overview

Your isoscapes are optimized and your study species/material are confidently assigned to origin, but now you're left with all these posterior probability surfaces! Let's explore some ways you can interpret and summarize your results.

## Setting up

```{r}
#| label: setup
library(tidyverse)
library(assignR)
library(terra)
library(rnaturalearth)
library(sf)
library(tidyterra)
library(viridis)
library(isocat)
```

::: {collapse="true"}
```{r}
unknown <- read.csv("Module3_assignment/unknown.csv")
Sr_iso_mean <- rast("Module3_assignment/rf_all_pred.tif")
Sr_iso_sd <- rast("Module3_assignment/rf_all_sd.tif")
Sr_iso <- c(Sr_iso_mean, Sr_iso_sd)
d2Hw_isoscape <- terra::rast("Module3_assignment/d2Hw_isoscape.tif")
ENM_Sept <- terra::rast("Module3_assignment/ENM_Sept.tif")
iso_vals  <- values(d2Hw_isoscape[[1]], na.rm = TRUE)
iso_range <- range(iso_vals)
unknown_dual <- unknown %>%
  mutate(in_range = d2H >= iso_range[1] & d2H <= iso_range[2]) %>%
  filter(in_range == T) %>%
  filter(Sample_ID != "RVcoll14B025") %>%
  dplyr::select(Sample_ID, Sr8786, d2H)
dual_isoscapes <- isoStack(Sr_iso, d2Hw_isoscape)
prior_probsurfaces <- pdRaster(r = dual_isoscapes, unknown = unknown_dual, prior = ENM_Sept)
eck <- crs(Sr_iso_mean)
WGS84 <- '+init=EPSG:4326'
sa <- ext(c(16.5, 35, -36, -22)) #(xmin, xmax, ymin, ymax
SA_outline <- rnaturalearth::ne_countries(
  country = c("South Africa", "Lesotho", "Eswatini"),returnclass = "sf") %>%
  sf::st_as_sf(., crs=WGS84) %>%
  st_crop(sa) %>% 
  sf::st_transform(crs=eck) 
unknown_dualxy<-unknown %>%
  mutate(in_range = d2H >= iso_range[1] & d2H <= iso_range[2]) %>%
  filter(in_range == T) %>%
  filter(Sample_ID != "RVcoll14B025") %>%
  st_as_sf(., coords=c("Lon","Lat"), crs=WGS84)%>% 
  st_transform(crs=eck)
```
:::

## Stacked maps

One classic way we could choose to deal with the large amount of information contained in a probability surface is to make the surface binary: areas of "low" and "high" probability.

```{r}
prior_binary <- assignR::qtlRaster(pdR = prior_probsurfaces, threshold = 0.333, thresholdType = "prob")
```

::: callout-tip
### Tip

It's become traditional to use a 2:1 odds ratio, but you can explore and optimize the odds ratio used (check out the assignR:QA function for inspiration)
:::
### map it!

A stacked map is the sum of the binary surfaces.

```{r}
#| message: false
#| warning: false
prior_stacked <- sum(prior_binary)

ggplot()+
  geom_spatraster(data=prior_stacked)+ # stacked map
  geom_spatvector(data=SA_outline,fill=NA,colour="black")+ # country outline
  geom_sf(data=unknown_dualxy,pch=1, col="red3",cex=2.5)+  # butterfly collection sites
  scale_fill_viridis(option="turbo", na.value = NA)+ #pretty colors
  theme_minimal()+
  theme(legend.position="bottom")+
  theme(legend.key.width=unit(2, "cm"))+
  guides(fill = guide_colourbar(direction = 'horizontal', ##transform legend
                                title= "what do you think the legend title should be?", ##rename default legend
                                title.position='top',
                                title.hjust=0.5,
                                ticks.colour='#f5f5f2',
                                ticks.linewidth=2,
                                barwidth = 15,
                                barheight = 0.75))  
```

::: {.think-pair-share}
### 🗺️ Think · Pair · Share: reading the stacked map

::: {.tps-steps}
::: {.tps-step}
**💭 Think** <span class="tps-time">3 min</span>

Your map has no legend title yet. On your own:

- What does each layer of `prior_binary` hold, and what happens when we `sum()` them?
- What is the smallest and largest value a cell could have?
- Write a legend title that would make sense to someone who hasn't done this workshop.
:::

::: {.tps-step}
**🤝 Pair** <span class="tps-time">3 min</span>

Before you look closely at your map, predict what it would look like if:

- 🎯 all the butterflies came from the same place
- 🌍 the butterflies came from many different places

:::

::: {.tps-step}
**📣 Share** <span class="tps-time">3 min</span>

Now look at your map. Which prediction is it closer to?

- Where are the hotspots, and how do they compare with the collection sites (red circles)?
- What can this map tell you, and what can't it tell you?
:::
:::
:::

::: {.callout-caution collapse="true"}
## Hint: check the numbers

```{r}
#| label: stack-scale
nlyr(prior_binary)                            # how many butterflies?
global(prior_binary, "sum", na.rm = TRUE)     # how many cells each butterfly selected
global(prior_stacked, range, na.rm = TRUE)        # range of the stacked map
```
:::

## Cluster analysis

For single isotope analyses, clustering with k-means etc works well. However, once you move to combining multiple geolocation tools, you may want to compare the spatial patterns of the probability surfaces directly.

### Spatial similarity with isocat

The package *isocat* has been specifically designed for this purpose! Let's check it out!

::: {.callout-tip icon="false"}
### More resources
-   Campbell, C. J., Fitzpatrick, M. C., Vander Zanden, H. B., & Nelson, D. M. (2020). Advancing interpretation of stable isotope assignment maps: Comparing and summarizing origins of known-provenance migratory bats. Animal Migration, 7(1), 27–41. <https://doi.org/10.1515/ami-2020-0004>
-   <https://github.com/cjcampbell/isocat>
:::

```{r}
#Applies pairwise comparisons of Schoener's D-metric between each layer in a spatraster to populate a similarity matrix.
prior_simmatrix <- isocat::surfaceSimilarityMatrix(prior_probsurfaces)
prior_simmatrix
```
```{r}
# wrapper for pvclust that applies hierarchical clustering analysis to similarity matrix
prior_cS <- isocat::clusterSimmatrix(simmatrix = prior_simmatrix)
plot(prior_cS)
```
```{r}
#choose a cutoff value
myheight <- 0.5
unknown_dual$cluster <- dendextend::cutree(prior_cS$hclust, h = myheight)
unknown_dual
```
For each cluster, you could make stacked maps or a map of the average probability
```{r}
# summary maps of each cluster
prior_meanSurfaces <- meanAggregateClusterProbability(indivIDs = unknown_dual$Sample_ID, 
  clusters = unknown_dual$cluster, 
  surfaces = prior_probsurfaces, 
  nClust = FALSE)

plot(prior_meanSurfaces) 
```

::: {.callout-tip icon="false"}
## 🧩 Food for thought

How else could you explore the probability surfaces? Was anything flagged in the exploratory data analysis? By date? By collection site? By sex?
:::

::: {.callout-tip icon="false"}
## 🧩 Other post-hoc analyses

There are many other follow up analyses for you to explore and consider for your own studies! Here are a few:

-   isotope-based distance metrics

-   direction

-   odds of originated from Region A vs Region B
:::

## Summary

::: {.callout-note appearance="simple"}
## Key takeaways

-   Interpreting probability surfaces confidently requires an in-depth understanding of
    -   the strengths and weaknesses of the underlying isoscapes
    -   the assumptions of the probability density function
    -   the ecology of your species
:::

Thank you so much for your attention; I hope you learned something new!

For comments, questions, or just to say hi, email Megan Reich at meganreich13\@gmail.com

```{r}
sessionInfo()
```
